Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-20 agujournaltemplate.tex
rewriting Intro 10%

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.

    % Phytoplankton is under threat
    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}. Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}. The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    % CARIACO is one such time series
    One such project is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    %%% mention climate impacts (AMO + ENSO) here!

@@

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
Variable Name & Units & Data source \\
\midrule
MEI v.2 & unitless & Climate Indices provided by NOAA \\
AMO & unitless & Climate Indices provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
Evaporation & NA & ERA5 model reanalysis: evaporation \\
SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

>>>>

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.

    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}. The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.

    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

@@

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.

@@

\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
\centering
\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
\toprule
Variable Name & Units & Mean & Range & Std. Dev. & Data source \\
\midrule
MEI v.2 & unitless & Climate Indices provided by NOAA \\
AMO & unitless & Climate Indices provided by NSF NCAR \\
Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
Evaporation & NA & ERA5 model reanalysis: evaporation \\
SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
Salinity & PSU & Niskin, top 100m \\
NO3 & µM & Niskin, top 100m \\
PO4 & µM & Niskin, top 100m \\
SiO4 & µM & Niskin, top 100m \\
\end{tabular}}
\label{tab:DataSources}
\end{table}

@@

- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}. % <- for this paper, check references to ENSO MEI v.2!!! This could be interesting!

### 2025-05-20 text-backup.tex
rewriting Intro 10%

>>>>

    % Phytoplankton is under threat
    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}. Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}. The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    %%%%%%%INTRO BACKUP REWRITE MAY "="%%%%%%%%

      % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.

    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}. The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.

    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% ADD Duplicate Points here:
    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.

    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."  
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.

    %% Gaps that We Adresse:
    % - More in-depth exploration of microscopy data & diversity time series
    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."

    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.

    % NOw we come in!
    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community. We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies. We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
